---
title: "Practical Machine Learning Project"
author: "Louis Jordan"
date: "Sunday, April 10, 2016"
output: html_document
---

### Executive Summary
Participants in the Human Activity Recognition (HAR) Project were asked to
perform various exercises correctly and incorrectly in 5 different ways.Using 
performance data collected from accelerometers fed by multiple quantified self 
movement devices, the goal of this project is to train a model that could be 
used to predict the manner in which the participants performed the exercises.

### Preliminaries
Set Seed for Reproducibility  
Load required libraries

```{r setup, include = FALSE}
      knitr::opts_chunk$set(cache = TRUE)
```


```{r seed, echo = TRUE, message = FALSE}
      set.seed(212061996)
      library(caret)
```

#### Load the Data Sets
```{r load, echo = TRUE, message = FALSE}
      trainChunk <- read.csv("pmlTrain.csv", header = TRUE, stringsAsFactors = FALSE, 
                       sep = ",", na.strings = c("NA", "", "#DIV/0!"))

      testChunk <- read.csv("pmlTest.csv", header = TRUE, stringsAsFactors = FALSE, 
                       sep = ",", na.strings = c("NA", "", "#DIV/0!"))

      trainChunk$classe <- as.factor(trainChunk$classe)
```

#### Preprocess (Examine & Clean) the Data
Remove missing values, irrelevant columns of data, and other items from the 
data set that do not contribute to the scope of the project
#### Examine the Data
```{r examine, echo = TRUE, message = FALSE}
      # summary(trainChunk)
      # str(trainChunk)
      dim(trainChunk)
```

#### Remove RowID Column
```{r removeRowID, echo = TRUE, message = FALSE}
      removeIDCol <- trainChunk[, -1]
      processedTrainChunk <- removeIDCol
      dim(processedTrainChunk)
```

#### Determine Missing Values
```{r findNAs, echo = TRUE, message = FALSE}
      NAs <- apply(processedTrainChunk, 2, 
                  function(x) {
                        sum(is.na(x))
                        }
                  )
```

#### Remove Missing Values
```{r removeNAs, echo = TRUE, message = FALSE}
      removeNAs <- processedTrainChunk[, which (NAs == 0)]
      processedTrainChunk <- removeNAs
      dim(processedTrainChunk)
```

#### Remove NZV Values
```{r removeNZVs, echo = TRUE, message = FALSE}
      removeNZV <- nearZeroVar(processedTrainChunk)
      processedTrainChunk <- processedTrainChunk[, -removeNZV]
      dim(processedTrainChunk)
```

#### Find & Remove Useless Predictors (features)
```{r removeUseless, echo = TRUE, message = FALSE}
      uselessPredictors <- grep("cvtd_timestamp|X|user_name|num_window", 
                                names (trainChunk))
      processedTrainChunk <- processedTrainChunk[, -uselessPredictors]
      dim(processedTrainChunk)
```

### Define Data Partitions
#### Create Training and Test Data Sets
```{r createDataSets, echo = TRUE, message = FALSE}
      inTrain <- createDataPartition(y = trainChunk$classe, p = 0.20, list = FALSE)
      trainSubset <- processedTrainChunk[inTrain, ]
      dim(trainSubset)
      # create test data for future use in cross validation
      testSubset <- processedTrainChunk[-inTrain, ] 
      dim(testSubset)
```

### Define & Fit the Model; Display Accuracy of the Model
#### Set Parameters of Train Control & Fit Using Random Forest Algorithm
```{r defineModel, echo = TRUE, message = FALSE}
      ctrl <- trainControl(method = "cv", number = 4, allowParallel = TRUE)
      modelFit <- train(trainSubset$classe ~ ., data = trainSubset, method = "rf", 
                prof = TRUE, trControl = ctrl)
      modelFit
      
      # Display the Accuracy of the Model
      resultsTR <- modelFit$results
      round(max(resultsTR$Accuracy), 4) * 100
```


### Process the Test Data
#### Load the test Data
```{r processTestData, echo = FALSE, message = FALSE}
      testChunk <- read.csv("pmlTest.csv", header = TRUE, stringsAsFactors = FALSE, 
                      sep = ",", na.strings = c("NA", "", "#DIV/0!"))     
```

#### Remove RowID Column
```{r remRowIDFromTest, echo = FALSE, message = FALSE}
      removeIDCol <- testChunk[, -1]
      processedTestChunk <- removeIDCol
      dim(processedTestChunk)
```

#### Find & Remove Missing Values
```{r remMissingFromTest, echo = FALSE, message = FALSE}
      NAs <- apply(processedTestChunk, 2, function(x) {sum(is.na(x))})
      removeNAs <- processedTestChunk[, which (NAs == 0)]
      processedTestChunk <- removeNAs
      dim(processedTestChunk)
```

#### Remove NZVs
```{r removeNZVsTest, echo = FALSE, message = FALSE}
      removeNZV <- nearZeroVar(processedTestChunk)
      processedTestChunk <- processedTestChunk[, -removeNZV]
      dim(processedTestChunk)
```

#### Remove Useless Predictors (features)
```{r remUseless, echo = FALSE, message = FALSE}
      uselessPredictors <- grep("cvtd_timestamp|X|user_name|num_window", 
                                names (testChunk))
      processedTestChunk <- processedTestChunk[, -uselessPredictors]
      dim(processedTestChunk)

      #testChunk$classe <- as.factor(testChunk$classe)
      tester <- testChunk[ , which(names(testChunk) %in% names(processedTrainChunk))]
```

### Predictions Using the Test Data
```{r predictions, echo = TRUE, message = FALSE}
      tcPrediction <- predict(modelFit, tester, type = "raw")
      tcPrediction
```


